Exploring probability distributions for bivariate temporal granularities

### 2019-11-07 main.Rmd
smart meter study update

Smart meters provide large quantities of measurements on energy usage for households across Australia, and indeed many parts of the world. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.

<!-- Existing approaches, why we need to drill down, or why we want to -- probability distributions -->
It is common to see aggregates of usage across households, total kwh used each half-hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.

One of the customer trial [@smart-meter] conducted as part of the Smart Grid Smart City (SGSC) project (2010-2014) in Newcastle, New South Wales and some parts of Sydney provides customer wise data on half-hourly energy usage and detailed information on appliance use, climate, retail and distributor product offers, and other related factors. It would be interesting to explore the energy consumption distribution for these customers and gain more insights on their energy behavior which are otherwise lost either due to aggregation or looking only at coarser temporal units. The idea here is to show how looking at the time across different granularities together can help identify different behavioral patterns. We look at the behavior of typical and extreme behaviors of 50 households from that trial.

@@

Figure \ref{fig:my-hd} shows the overlay quantile plot of energy consumption of 50 households. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. Each facet represents a month and energy consumption across each hours of the day is shown inside each facet. It can be observed that the median is very close (and hence not distinctly visible) to the lower boundaries of all the other bands implying energy consumption for these households are extremely left skewed.75% of the households has distinct higher evening peaks compared to morning peaks, especially in the winter months. The top 5% households behave very differently than even the top 10% households in terms of total energy consumption across hours of the day.

>>>>

Smart meters provide large quantities of measurements on energy usage for households across Australia. One of the customer trial [@smart-meter] conducted as part of the Smart Grid Smart City (SGSC) project (2010-2014) in Newcastle, New South Wales and some parts of Sydney provides customer wise data on half-hourly energy usage and detailed information on appliance use, climate, retail and distributor product offers, and other related factors. It would be interesting to explore the energy consumption distribution for these customers and gain more insights on their energy behavior which are otherwise lost either due to aggregation or looking only at coarser temporal units. The idea here is to show how looking at the time across different granularities together can help identify different behavioral patterns and identify the extreme and regular households amongst these 50 households.

<!-- typical and extreme behaviors of 50 households from that trial.  -->

<!-- , and indeed many parts of the world. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.  -->

<!-- <!-- Existing approaches, why we need to drill down, or why we want to -- probability distributions --> -->
<!-- It is common to see aggregates of usage across households, total kwh used each half-hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.  -->

@@

In Figure \ref{fig:day_fortnight} and \ref{fig:day_fortnight_agg} , the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy beahvior of these two customers. In Figure \ref{fig:day_fortnight_agg} it can be seen that for most days of the fortnight, the second household has much less consumption than the first one. However, Figure \ref{fig:day_fortnight} adds to the beahvior by looking at the distribution. If we consider letter value F as a regular behavior and letter values beyond F as not-so-regular behavior, we can conclude that the regular behavior of the first household is more stable than the second household. However, the distribution of tail of the first household is more variable, observed through distinct letter values, implying that their not-so-regular behavior is quite extreme. This shows, how looking at the distribution of the dependent variable can throw more light on the energy behavior of the customers, which are lost using aggregate or summary statistics.

```{r day_fortnight_agg}
library(lvplot)
library(ggplot2)
library(dplyr)
library(tibble)
library(gravitas)
library(tidyr)
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704, 10017936)) %>% 
#   create_gran("day_fortnight") %>%
#   as_tibble() %>% 
#   group_by(customer_id, day_fortnight) %>% 
#   summarise(median= log(median(general_supply_kwh)),
#             total = log(sum(general_supply_kwh))) %>% 
#   gather("daily_consumption", "value", - c(customer_id, day_fortnight)) %>% 
#   ggplot2::ggplot(aes(
#     x = as.factor(day_fortnight),
#     y = value, 
#     color = daily_consumption,
#     group = daily_consumption)) +
#   geom_line() +
#   xlab("day_fortnight") +
#   ylab("general_supply_kwh") + 
#   facet_wrap(~customer_id) + 
#   theme_bw()

smart_meter10 %>% 
  filter(customer_id %in% c(10006704, 10017936)) %>% 
  create_gran("day_fortnight") %>%
  as_tibble() %>% 
  group_by(customer_id, day_fortnight) %>% 
  summarise(median_consumption= median(general_supply_kwh)) %>%
  ggplot2::ggplot(aes(
    x = as.factor(day_fortnight),
    y = median_consumption,
    color =  customer_id,
    group = customer_id)) +
  geom_line() +
  xlab("day_fortnight") +
  ylab("general_supply_kwh") + 
  theme_bw() 
```

```{r day_fortnight}
library(lvplot)
library(ggplot2)
library(dplyr)
library(tibble)
library(gravitas)
smart_meter10 %>%
  filter(customer_id %in% c(10006704, 10017936)) %>% 
  create_gran("day_fortnight") %>%
  ggplot2::ggplot(aes(
    x = as.factor(day_fortnight),
    y = general_supply_kwh)) +
  xlab("day_fortnight") +
  geom_lv(
    outlier.colour = "red",
    aes(fill = ..LV..),
    k = 5) +
  facet_wrap(~customer_id) + 
  scale_fill_lv() +
  theme_bw() +
  scale_y_sqrt() 
```

<!-- Figure \autoref{fig:my-hd} shows the overlay quantile plot of energy consumption of 50 households. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. Each facet represents a month and energy consumption across each hours of the day is shown inside each facet. It can be observed that the median is very close (and hence not distinctly visible) to the lower boundaries of all the other bands implying energy consumption for these households are extremely left skewed.75% of the households has distinct higher evening peaks compared to morning peaks, especially in the winter months. The top 5% households behave very differently than even the top 10% households in terms of total energy consumption across hours of the day.  -->

While trying to explore the energy behavior of these customers systematically, the first thing we should have at our disposal for examining periodicities of energy behavior across time granularities is to know the number of time granularities we can look at exhaustively. If we consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, fortnight, month, quarter, semester, year), the following time granularities can be considered for this analysis.

```{r}
library(tsibble)
interval(smart_meter10)
smart_meter10 %>% search_gran()
```

The interval of this tsibble is 30 minutes, and hence the default for search_gran in this case, provides temporal granularities ranging from half-hour to year. If these options are considered too many, the default options can be modified to limit the possibilities. For example, the most coarse temporal unit can be set to be a “month”.

```{r}
smart_meter10 %>% 
  search_gran(highest_unit = "month")
```

This looks better. However, some intermediate temporal units might not be pertinent to the analysis and we might want to remove them from the list of granularities.

```{r search_gran_limit2}

smart_meter10 %>% search_gran(highest_unit = "month",
                              filter_out = c("hhour", "fortnight")
)
```

Now that we have the list of granularities to look at, we can visualize the distribution. From the search list, we found that we can look at six granularities, that amounts to analyzing six graphics. However, what happens if we want to see the distribution of energy across two granularities at a time? This is equivalent to looking at the distribution of energy consumption across one granularity conditional on another one. One way can be to plot one of the granularities on the x-axis and another on the facet. Different perspectives of the data can be derived depending on where the granularities are placed.

So, what is the number of pairs of granularities we can look at? It is equivalent to taking 2 granularities from 6, which essentially means we need to examine 30 plots. The good news is, not all time granularities can be plotted together and we do not have to analyze so many plots!

Harmony/clash can be identified to considerably reduce the number of visualizations that can aid exploratory analysis.

```{r is_harmony, echo= TRUE}
smart_meter10 %>% 
  is_harmony(gran1 = "hour_day", 
             gran2 = "day_week")

smart_meter10 %>%
  is_harmony(gran1 = "hour_day", 
             gran2 = "day_week", 
             facet_h = 14)

smart_meter10 %>% 
  is_harmony(gran1 = "day_month",
             gran2 = "week_month")
```

Let us now look at all the harmonies that we can examine. Fortunately, we are left with only 13 out of 30 visualizations.

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
```

In \ref{fig:bothcust} We visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across wknd_wday (facet) and hour-day (x-axis) for the same two households. For the second household, outliers are less prominent implying their regular behavior is more stable. For the first household, energy behavior is not significantly different between weekdays and weekends. For the second household, median energy consumption for the early morning hours is extremely high for weekends compared to weekdays.

 For the second household, outliers are less prominent implying their regular behavior is more stable. For the first household, energy behavior is not significantly different between weekdays and weekends. For the second household, median energy consumption for the early morning hours is extremely high for weekends compared to weekdays.

```{r}
cust1 <- smart_meter10 %>% 
  filter(customer_id %in% c(10006704)) %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "boxplot") +
  scale_y_sqrt() +
  ggtitle("Energy consumption distribution for customer id: 10006704")

cust2 <- smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "boxplot") +
  scale_y_sqrt() + 
  ggtitle("Energy consumption distribution for customer id: 10017936")
```

```{r bothcust}
cust1
cust2
```

Now, we would like to see how these customers' behavior relate to the rest of the 
50 households across these two measures -

- if the regular behavior is regular and extreme behavior extreme or vice versa
- if the weekend and weekday behavior are different for most of these households.

Figure \ref{fig:wknd_wday50} shows the overlay quantile plot of energy consumption of 10 households. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile.  It can be observed that the median is very closer to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. Weekend and weekday behavior is not very different for these ten customers, implying the first customer is more typical than the second customer when compared to these ten customers.

```{r wknd_wday50}
smart_meter10 %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = TRUE, 
            quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt()
```
